\chapter{\abstractname}
Recent crises, natural disasters, and large-scale protests have highlighted the significant reliance of people on online social networks and instant messaging platforms for communication.
During crises, consolidated communication systems can become strained, partially or wholly destroyed, censored, or otherwise monitored.
Permissionless distributed overlays are alternatives to consolidated communication realizations.
System designers consider these overlays to be more resilient, since they distribute power and content control among many equal peers.
These overlays introduce a massive performance overhead, as they rely on gossip and constant chatter between peers to disseminate the correct system state by reaching consensus, through many rounds of communication.
This performance overheaf poses a risk to usablity of such systems in disaster zones, as peer devices often have limited connectivity and power supply, which together with the gossip protocol, can severely limit the peers' ability to communicate.
In this paper, we propose a system architecture for unstructured peer-to-peer networks that achieves grassroots constitutional consensus based on efficient geometrical overlay design that trade achieve a higher throughput by using more edge computing.
We use this system architecture to evaluate a decentralized social network deployed on smartphones using BLE connectivity, where users have total control of their data and privacy.




\makeatletter
\ifthenelse{\pdf@strcmp{\languagename}{english}=0}
{\renewcommand{\abstractname}{Kurzfassung}}
{\renewcommand{\abstractname}{Abstract}}
\makeatother

\chapter{\abstractname}

%TODO: Abstract in other language
\begin{otherlanguage}{ngerman} % TODO: select other language, either ngerman or english !

\end{otherlanguage}


% Undo the name switch
\makeatletter
\ifthenelse{\pdf@strcmp{\languagename}{english}=0}
{\renewcommand{\abstractname}{Abstract}}
{\renewcommand{\abstractname}{Kurzfassung}}
\makeatother