\chapter{State-of-the-Art}
\label{chap:sota}

The State-of-the-Art section presents the papers that lay at the foundation at this research work, and positions it adverseraly. The sections in this chapter range will provide an overview over the communication mediums that we considered for this work: Bluetooth Classic, Bluetooth Low Energy, Bluetooth Mesh, Wi-Fi ad-hoc and Direct, and others. Next, this section continues with elaborating the networking assumptions that make this work possible, primarily discussing the difference between traditional packet-switching and the Store-Carry-Forward paradigm, that enables Delay-Tolerant Networking, while also presenting similar systems that have been developed using this paradigm. Followingly, we present main approaches that enable packet propagation in Delay Tolerant Networks (DTNs), explain each one of them, and show which one we used. Then, we present leading Decentralized Social Networks, and discuss whether they could have been used in our scenarios, showing that they actually cannot, and finally present the pinnacle: Grassroots Social Networks. Grassroots Social Networks lay at the core of this research work, and the networking layer presented here is a core work that lays at its foundation.

\section{Proximate Peer-to-peer Communication Technologies}
\label{sec:proximate-p2p-protocols}
To create the peer-to-peer communication platform, different technologies were evaluated based on the following criteria:
\begin{itemize}
    \item \textbf{C1} Spread of the protocol: As a user-facing tool, the communication medium had to be present in virtually all smartphones used today; thus, technologies that are supported by a handful of flagship smartphones, while being an interesting point for future work, are out of the question for this work.
    \item \textbf{C2} Low power: the communication protocol should not use a high amount of power to propagate packets
    \item \textbf{C3} Target: the communication protocol should be able to carry text messages efficiently, so a text message should be received within a short time in close range (e.g. 200KB in 1 second). 
    \item \textbf{C4} Mobility: protocols neede to support a high node mobility stemming from the mobility patterns of the users
    \item \textbf{C5} Infrastructure: there can be no external infrastructure necessary for the normal operation of the technology
    \item \textbf{C6} The protocol should be operational even when the app is backgrounded
\end{itemize}
\subsection{Wi-Fi}
To relay information directly between peer smartphones, the omnipresent Wi-Fi protocol was first investigated. Propelled by the Wi-Fi Alliance, WLAN is the wireless 802.11 protocol which enables most of the world's computers to communicate. Together with cellular-based mobile networks like 5G, home and commercial routers provide wireless radio access networks for mobile devices to connect to the Internet backbone through a multitude of devices: antennas (Radio Units), computers for signal processing (Distributed Units), base stations (Centralized Units), and more~\cite{5g}.
For the standard Internet-backbone to function, all the aforementioned infrastructure units need to have a constant stream of power, an unblocked radio spectrum, and operational staff. In emergency scenarios such as natural disasters, wars, or politically-motivated violence against the freedom of speech through censorship, media control, or monitoring, the aforementioned infrastructure might not be available for usage.
As alternatives to the classical Wi-Fi WLAN technology to facilitate DTNs, we evaluated the following p2p communication media based on the aforementioned common criteria: Wifi Adhoc, Wifi Direct, Wifi Aware, Bluetooth BDR, Bluetooth Low Energy.
The Wi-Fi Ad-Hoc mode was the first design to adapt the WLAN standard to direct peer-to-peer wireless communication.
However, it suffered from several notable architectural shortcomings, including inefficient power management, a lack of robust security mechanisms, and the requirement for devices to manually decide whether to act as the access point or a client~\cite{2012IEEE80211}.
In the Wi-Fi Ad-Hoc mode, idle smartphones suffer from battery drain due to a lack of distributed power saving mode~\cite{Friedman2013}. However, actively communicating smartphones had the same power cost as using a standard access point, and comparing to Bluetooth, Wi-Fi Ad-Hoc still has a vastly superior throughput rate, which is optimal for larger file transfers.
In our context it is still not as useful due to \textbf{C3}: the transfer size is minimal, and we optimize for efficient battery usage through \textbf{C2}.
Wi-Fi Ad-Hoc was eventually replaced by Wi-Fi Direct, which modernized peer-to-peer connectivity by adding better security and an autonomous negotiation protocol that allows connecting devices to automatically determine which one takes over the role of the access point~\cite{CampsMur2013,Strobel2024}.
However, Wi-Fi Direct has two main limitations: it does not support device discovery in the background if the utilizing application is stopped or not actively running in the foreground; another significant constraint of Wi-Fi Direct on modern mobile platforms is that it is not supported by iOS devices, as Apple developed a competing standard, AirDrop~\cite{airdrop}.
The Android equivalent of AirDrop, QuickShare, is developed cooperatively by Samsung and Google, and since November 2025 supports platform interoperability between Android and iOS, and has had support transfers between Androids, Windows computers, and Chromebooks before.
While similarly providing superior data rates, QuickShare, as it uses a combination of Wi-Fi Direct and Bluetooth, is still not usable for this work, as its core lies with Wi-Fi Direct, which requires more power and devices to have the using app constantly foregrounded.
Wi-Fi Aware (also known as Neighbor Awareness Networking / NAN) overcomes the background limitation of Wi-Fi Direct by utilizing a highly power-efficient \textit{discover-before-connect} paradigm.
Unlike Wi-Fi Direct, Wi-Fi Aware allows devices to automatically organize into synchronized local clusters and use a publish-and-subscribe mechanism to discover nearby services and exchange lightweight messages in the background, all without establishing a formal network connection.
Because it operates without maintaining an active connection state, it has a significantly lower power and memory footprint.
If devices need to share large amounts of data, they can seamlessly upgrade their discovery session into a fully bi-directional, high-throughput network connection.
Positioning itself as the most appropriate for our use-case, Wi-Fi Aware seems like a promising candidate; however, it was primarly intended for IoT devices~\cite{Saloni2016}, support for it on iOS was only recently introduced~\cite{WWDC25}, and thus it was deemed not widely supported enough.
Early research into ad hoc and delay-tolerant networks frequently relied on legacy 802.11b and 802.11g standards, which were highly capable for active data transfers but suffered from severe battery drain when devices were left idle~\cite{Ott2006}.
In later standard, 802.11n has introduced a significantly higher throughput and a 17\% improvement in energy efficiency over previous standards~\cite{Friedman2013}.

\subsection{Alternatives}
With Wi-Fi standards being exhaused for picking the base communication protocol, we evaluated other protocols such as Zigbee and LoRA, but very quickly forsaken due to lacking support for direct communication between mobile phones: Zigbee, LoRA, and 6LowPAN are all intended for small IoT devices to communicate among themselves or with centralized servers~\cite{Saloni2016}.

\subsection{Bluetooth}
A technology that is ubiquotous to all smartphones that have been released in the last 10 years is Bluetooth. Being standardized as Bluetooth 1.0 in 1999 by the Bluetooth Special Interest Group (SIG), Bluetooth was introduced to mobile phones by Ericsson in 2001 with the release of the T36, which allowed phones to replace bulky cables transfers with a wireless connection.
Since then, Bluetooth has seen four major version bumps:
\begin{itemize}
    \item Bluetooth 2.0 offered in comparison to Wi-Fi a much lower energy consumption for transferring large files, and higher resilience to signal interference from nearby transmissions.
    \item Bluetooth 3.0 was introduced as an upgraded specification designed to improve upon the throughput limitations of Bluetooth 2.x, doubling both send/receive throughput speeds to almost 200 KB/s, while the increasing the overall power consumption~\cite{Friedman2013}.
    \item Bluetooth 4.0 represented an elementary shift in Bluetooth technology, with the focus shifting to optimizing for ultra-low power consumption: Bluetooth Low Energy, or BLE, was introduced in this version. BLE presents until today the backbone for most modern smartphone p2p applications, due to major power-saving capabilities: because BLE supports background advertising and discovery without requiring apps to be actively foregrounded, BLE is the primary transport mechanism for decentralized networks and emergency systems like firechat, Briar, Corona-Warn-App, Twimight, BitChat, and more~\cite{firechat,briar,CoronaWarnApp,bitchat}.
\end{itemize}

BLE is characterized by a dual-role split: advertising, which is carried out by the peripheral device, and scanning, which is done by the central device.
The central device controls the connection, while the peripheral provides the service.
Advertising happens on a set window, in which the peripheral sends out advertisements on an interval.
The advertisements contain a customizable device unique identifier, as well as custom manufacturer-specific data.
The BLE advertised services stem from the underlying General Attribute Protocol, or GATT.
The GATT protocol is a hierarchial protocol, where the highest level is GAP: General Attribute Profile. A GAP groups GATT services together, to offer a specific function, e.g. a heart rate monitor, or a grassroots networking profile.
Furthermore, GATT services group a set of characteristics, which are used for interfacing with data, and controlling settings pertaining to the service.
Each of the characteristic is identified by a UUID, a value, and has different read/write permissions; the data-flow between the peripheral and the central is bidirectional, with the peripheral being able to notify the central of any changes to a characteristic value as it changes, or central can request to read the current value of a characteristic, and more~\cite{BLE-guide}.
To see the peripheral's advertisements, the central device also has to scan during the same time: the scanning is also limited to a window, during which the central listens to all incoming advertisements from surrounding devices.
To discover GATT services, a device first must perform a connection to one such a device, which potentially includes pairing - depending on the permissions the device's GATT server defined.   
BLE presents advantages not only in its conservative use of battery power, but also in its security, and usage model. When an app registers a GATT serveer with the BLE adapter, the mobile phone controls it natively, and the app can remain backgrounded or on standby, while the OS continues to scan and advertise for services.
This allows for both a heightened security model due to the protocol abstraction by the operation system, as well as the grouping together of functionalities from different apps to a central native coordinator.
Providing a throughput of about 230 kbps at a low power usage~\cite{Jacopo2017}, BLE was able to answer all out criteria C1-C5.
Due to the aforementioned benefits of the protocol, we thus chose BLE as the application's main communication protocol.
\newcommand{\pmark}{$\sim$}    % partial / conditional

\begin{table}[htbp]
    \centering
    \caption{Evaluation of p2p communication technologies against the selection criteria.}
    \label{tab:p2p-criteria}
    \begin{tabular}{@{}l*{5}{P{1.6cm}}@{}}
        \toprule
        \thead[l]{Technology} & \thead{Spread} & \thead{Low\\power} & \thead{Target\\throughput} & \thead{Mobility} & \thead{No infra-\\structure} \\
        \midrule
        Wi-Fi Ad-Hoc          &  &  &  &  &  \\
        Wi-Fi Direct          &  &  &  &  &  \\
        Wi-Fi Aware           &  &  &  &  &  \\
        Bluetooth Classic     &  &  &  &  &  \\
        Bluetooth Low Energy  &  &  &  &  &  \\
        \bottomrule
    \end{tabular}
    \\[0.5em]
    {\footnotesize \cmark{}~fully met \quad \pmark{}~partially met \quad \xmark{}~not met}
\end{table}

\section{Remote Peer-to-peer Communication Technologies}
\label{sec:remote-p2p-protocols}
After discussing the choice for a proximate communication protocol, this section continues with elaborating on how different protocols were evaluated to facilitate long-distance communication.
The protocols that were evaluated in this section include UDP, TCP, HTTPS, QUIC, and UDX.

\subsection{Application Protocols: HTTP, HTTPS, QUIC}
HTTP and HTTPS are the first protocols that we evaluated, as they lie at the base of every modern web applications, and have been tried-and-tested by many applications.
Most web servers started by using HTTP, and in a standard web application scenario with a typical client and server having continuous end-to-end connectivity it would be optimal; however to deploy it in a DTN environment where connectivity is intermittent at best, multiple auxillary systems would need to be used in order to provide hop-by-hop connectivity. As our goal is to provide connectivity with the least amount of external infrastructure, HTTP and its secure variant HTTP over TLS/TLS (HTTPS) require many subsystems that cannot be guaranteed to be freely available in our scenarios.
Additionally, HTTPS requires having one constantly listening TCP socket, plus one live TLS connection per peer, each with crypto state. Android and iOS tend to aggresively kill backgrounded applications and sockets. Recovering the socket connections would cost a whole TCP handshake, and a TLS handshake, which would cost multiple round trips per peer.
This shows that recovery for TCP-and-TLS-based protocols is a heavy operation that is very expernsive in terms of time-to-first-message after start.
QUIC, the transport standardized underneath HTTP/3, was designed to overcome two structural limitations of the TCP+TLS stack: its serialized handshakes and head-of-line blocking.
Built on top of UDP, QUIC folds the transport and cryptographic handshakes into a single exchange, establishing an encrypted connection in one round trip, and in zero round trips when resuming a previously cached session; it further multiplexes independent streams over one connection, so that a lost packet on one stream does not stall the others~\cite{Langley2017}.
However, QUIC remains connection-oriented: it maintains per-connection state, negotiates a full TLS~1.3 handshake on every cold start, and, like TCP, must re-establish that state whenever the underlying socket is torn down.
In a mobile setting where the operating system aggressively reclaims backgrounded sockets, this coupling of transport and cryptographic state reintroduces the very time-to-first-message penalty we sought to avoid.
The first protocol that we consideres for that, which already implements all of the above, is QUIC, or Quick UDP Network Connections~/cite{Langley2017}.
QUIC was designed with a mobile user in mind, supporting seamless network migration through the usage of Connection IDs to route and identify connections, rather than a source/destination IP:port + protocol 5-tuple combination; multiplexing multiple lightweight streams within a single connection allows to avoid the Head-of-Line Blocking issue, which causes queues of packets to buffer and be held up until the first packet is fully processed, severely limiting the performance, and is further exacerbated through out-of-order delivery; finally, in an effort to make common pairings more efficient, QUIC combines the transport connection establishment handshake with a cryptographic handshake to offer 0-RTT connectino setup, allowing data to be sent directly on the first message to known peers.
Since QUIC combines the cryptographic layer (normally TLS) with the network layer, it requires continuous cryptographic processing, which reportedly cost up to 3.5 times more CPU than normal TCP using TLS~\cite{Langley2017}. This computational exertion quickly drains mobile devices, since they are highly battery-constrained, and their CPU is much weaker than the desktop computers QUIC was designed for. Additionally, one of the main advantages of QUIC is the 0-RTT handshake data delivery, which depends on cached address tokens. These cached address tokens often become invalidated when mobile phones roam, dropping the success rate significantly~\cite{Langley2017}.
However, precisely because it combines so many things that UDP normally does not offer, QUIC became a monolith of a protocol that is mostly designed for stable end-to-end connections such as traditional HTTP (HTTP/3) client-server connections. 

\subsection{Transport-level Protocols: TCP, UDP}
The elementary transport protocols that enable QUIC and HTTPS are UDP and TCP respectively: to decouple cryptography from connection maintenance and application state, we decided to focus on the transport protocols that enable application communication.
Providing congestion-controlled, reliable, ordered, error-checked streams, TCP, the core transport protocol behind the IP protocol stack, was the first protocol that we investigated.
As a congestion-controlled transport protocol, TCP assumes that all transmission errors in its delivery path stem from link congestion in the network; that means, that for intermittent connectivity outages, TCP will engage in congestion-avoidance measures, which dramatically reduce the possible data rate on the link temporaily~\cite{RFC5681}.
Moreover, increasing the number of end-to-end hops in the network connection yields rapidly deterioratating performance with the increase of hops~\cite{Ott2010}.
Additionally, the usage of Network Address Translation (NAT) in order to assign multiple hosts to the same IPv4 address in a network makes it even more difficult to establish a p2p TCP connection without using any external proxies, as the NAT mapping is dependent on a 5-tuple: both peers IP addresses, ports, and protocol~\cite{Keidar2026}.
Because TCP maintains a strict connection state, the connection will break entirely any element in the 5-tuple formed by the underlying network changes.
Every time a mobile device roams, such as switching from a Wi-Fi network to cellular data, or changing its network attachment point, its IP address changes, forcing the TCP connection to be completely re-established from scratch.
To overcome the obstacles caused by node mobility and the ensuing change in Internet addresses, we have decided to focus on the UDP protocol as we expect frequent address changes due to our mobility model.

\subsection{UDX}
\label{sec:udx}
The transport-layer protocol that was chosen for the system, UDP, is a lightweight, data-oriented network protocol that allows the users to send datagrams without establishing a session or connection between each other. It prioritizes simplicity, incarnate in a lightweight header size, over error-correcting code, reliability, or packet-ordering which are all present in TCP, which makes it ideal for sending lightweight messages as it wastes no time establishing a reliable connection as its counterpart TCP.
UDP or UDP-based protocols are often used in time-critical applications like online gaming, streaming, and other p2p applications.
Due to the focus on simplicity, the resulting trade-off yields a transport-layer that, had it based itself directly on UDP, would have needed to implement error-correction codes, deal with network-level packet-loss, congestion control and flow control on active links. In addition to the transport-level control mechanisms, an encryption layer needs to be implemented on-top, in order to guarantee no adversary could read messages that were not intended for them.
The transport we ultimately selected for remote peer-to-peer links is UDX, the reliable stream transport that underpins the Holepunch/Pears networking stack~\cite{udx}.
UDX provides reliable, in-order, multiplexed, and congestion-controlled streams over plain UDP, much like QUIC, but with one deliberate design difference: it performs no handshake and no encryption of its own.
As the reference implementation phrases it, UDX has ``no handshakes, no encryption, no features''~\cite{udx}, which is precisely what makes it inexpensive to (re-)establish.
Because UDX is connectionless, a stream is set up by each side independently creating a stream with a locally chosen identifier and connecting to the remote peer using the stream identifier and the remote socket address. Neither peer waits for an acknowledgement before it may send.
The very first packet can therefore already carry application payload, making the transport itself effectively \emph{zero round trip}.
Encryption and authentication are provided by separate layers: the Noise suite-based~\cite{noise} secure public key exchanged is performed after the application-level identity exchange happens, with session keys derived from the identity public keys through a birational map~\cite{Zafar2025}.
Crucially, this cryptographic handshake is not paid for directly in the transport layer, which makes it very simple to establish such a connection.

\section{Opportunistic Networking Systems}
\label{sec:oppnet}
From the communication protocols we elaborate to research other existing opportunistic networking systems.
MANETs, who predate OppNets, seemed promising to enable ad-hoc communcication between mobile devices, requiring devices to forward messages for others in addition to allowing the devices themselves to communicate.
Talk about different routing strategies in MANETs?
One of the leading examples of MANETs that are still developed to this day, is the BATMAN-adv protocol (Better Approach To Mobile Ad-hoc Networking), by the Freifunk Gesellschaft.
In BATMAN, nodes constantly broadcast a hello message to their peers, and peers that receive those messages propagate those further. Peers measure the proportion of hello messages that they receive from a node to estimate their link quality, and broadcast that to their peers as well. Peers then re-broadcast the propagation from the peer with the highest ratio link quality, and thus a routing infrastructure is built iteratively.
Opportunistic Networks (OppNets) are a subsystem of DTNs, which are an evolution of legacy MANET systems.
In MANET systems there has to be a fully-connected end-to-end path before a message can be sent.
OppNets are an instance of the Store-Carry-Forward paradigm, where the obstacle that prevents MANETs from being used, is used as an advantage - node mobility. In Packet-Switched Networks, an alias to OppNets, peers frequently change their network attachment point due to physical movements. In those networks, peers carry messages with them as they move until the opportunity arises to forward the message to another peer~\cite{Boldrini2014}.
In OppNets, the emphasis lies on the opportunistic establishment of links between peers. Every possible participating user is henceforth referred to as a peer in the OppNet, and the links between them can be establish opportunistically: that means a peer randomally passing another peer will try to establish a (secure) connection to it.

(MANET BATMAN) vs. Twimight, Emergency Broadcast, COVID contact tracing, Wifi-Opp


\section{Routing in Opportunistic Networks}
\label{sec:opp-routing}
Flooding, Spray-and-Pray, PRoPHET

\section{Decentralized Social Networks}
\label{sec:dsn}
Bluesky, Mastodon, Briar, Nostr, SBB (Manyverse), Bitchat

\section{Grassroots Social Networks}
\label{sec:gsn}
Shapiro